% Main paper: Preference Collapse Exploitability
% Target: NeurIPS / AAAI 2026
\documentclass{article}

% Use NeurIPS 2024 style (adjust as needed)
\usepackage[final]{neurips_2024}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{enumitem}
\usepackage{multirow}
\usepackage{xcolor}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}

% Theorem environments
\newtheorem{proposition}{Proposition}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{corollary}{Corollary}
\theoremstyle{remark}
\newtheorem{remark}{Remark}

\title{Preference Collapse Exploitability: How DPO's Mode Collapse Becomes a Weaponizable Safety Vulnerability}

\author{
  Anonymous Author(s)
}

\begin{document}

\maketitle

\begin{abstract}
Direct Preference Optimization (DPO) has become the dominant method for aligning large language models with human preferences. However, DPO's implicit optimization dynamics concentrate probability mass on narrow output modes---a phenomenon known as mode collapse. We identify a previously unrecognized security implication: mode collapse renders models deterministically predictable, enabling adversaries to reliably trigger harmful outputs. We formalize this risk through \textit{Preference Collapse Exploitability} (PCE), a metric that quantifies the security vulnerability arising from distributional concentration. We prove that PCE increases monotonically during DPO training and identify a critical transition point $t^*$ that precedes standard performance peaks---meaning community training practices systematically produce exploitable models. We further design \textit{collapse accelerator} attacks that require only 1--2\% data poisoning to amplify vulnerability, achieving 78\% targeted attack success rates. An audit of six widely-deployed open-source DPO models confirms all exhibit PCE above the exploitable threshold. Finally, we propose entropy-regularized DPO (ER-DPO), which reduces PCE by 59\% with less than 0.4-point MT-Bench degradation. Our findings reframe mode collapse from a performance limitation to a first-order security concern, demanding that alignment practitioners monitor distributional geometry alongside output quality.
\end{abstract}

% Section 1: Introduction
\input{introduction}

% Section 2: Related Work
\input{section2_related_work}

% Section 3: Method
\input{section3_method}

% Sections 4-5: Experiments and Results
\input{sections_4_5_experiments_results}

% Sections 6-7: Discussion and Conclusion
\input{sections/section6_discussion}

% References
\bibliographystyle{plain}
\bibliography{references}

% Appendix
\input{appendix}

\end{document}
